### 2024-06-26 basic-statistics.qmd
20240626 - scores, scales, and regression

@@

Regression with one predictor takes the form of @eq-regression:

$$
y = \beta_0 + \beta_1x_1 + \epsilon
$$ {#eq-regression}

where $y$ is the [outcome variable](#sec-correlationalStudy), $\beta_0$ is the intercept, $\beta_1$ is the slope, $x_1$ is the [predictor variable](#sec-correlationalStudy), and $\epsilon$ is the error term.

```{r}
#| include: false
#| eval: false

set.seed(52242)
regression <- data.frame(outcome = rnorm(40, mean = 5, sd = 2))

regression$predictor <- complement(regression$outcome, .5)
regression$predictor <- regression$predictor + abs(min(regression$predictor))

lm(
  outcome ~ predictor,
  data = regression)

ggplot2::ggplot(
  data = regression,
  ggplot2::aes(
    x = predictor,
    y = outcome,
  )
) +
  ggplot2::geom_point() +
  ggplot2::geom_smooth(
    method = "lm",
    linewidth = 2,
    se = FALSE,
    fullrange = TRUE) +
  ggplot2::scale_x_continuous(
    lim = c(0,8),
    breaks = seq(from = 0, to = 8, by = 2),
    expand = c(0,0)
  ) +
  ggplot2::scale_y_continuous(
    lim = c(0,8),
    breaks = seq(from = 0, to = 8, by = 2),
    expand = c(0,0)
  ) +
  ggplot2::labs(
    x = "Predictor Variable",
    y = "Outcome Variable",
    title = "Regression Best-Fit Line"
  ) +
  ggplot2::theme_classic(
    base_size = 16) +
  ggplot2::theme(legend.title = element_blank())

ggsave("./images/regression.pdf", width = 6, height = 6)
```

A regression line is depicted in @fig-regression.

::: {#fig-regression}
![](images/regression.png)

A Regression Best-Fit Line.
:::

Regression with multiple predictors—i.e., multiple regression—takes the form of @eq-multipleRegression:

$$
y = \beta_0 + \beta_1x_1 + \beta_2x_2 + ... + \beta_px_p + \epsilon
$$ {#eq-multipleRegression}

where $p$ is the number of [predictor variables](#sec-correlationalStudy).

>>>>

## Scores and Scales {#sec-scoresAndScales}

There are many different types of scores and scales.
This book focuses on [raw scores](#sec-rawScores) and [*z*-scores](#sec-zScores).
For information on other scores and scales, including percentile ranks, *T*-scores, standard scores, scaled scores, and stanine scores, see here: <https://isaactpetersen.github.io/Principles-Psychological-Assessment/scoresScales.html#scoreTransformation> [@PetersenPrinciplesPsychAssessment].

### Raw Scores {#sec-rawScores}

*Raw scores* are the original data on the original metric.
Thus, raw scores are considered *unstandardized*.
For example, raw scores that represent the players' age may range from 20 to 40.
Raw scores depend on the construct and unit; thus raw scores may not be comparable across variables.

### *z* Scores {#sec-zScores}

*z* scores have a mean of zero and a standard deviation of one.
*z* scores are frequently used to render scores across variables more comparable.
Thus, *z* scores are considered a form of a *standardized* score.

*z* scores are calculated using @eq-zScore:

$$
z = \frac{X - \bar{X}}{\sigma}
$$ {#eq-zScore}

where $X$ is the observed score, $\bar{X}$ is the mean observed score, and $\sigma$ is the standard deviation of the observed scores.

You can easily convert a variable to a *z* score using the `scale()` function:

```{r}
#| eval: false

scale(variable)
```

With a standard normal curve, 68% of scores fall within one standard deviation of the mean.
95% of scores fall within two standard deviations of the mean.
99.7% of scores fall within three standard deviations of the mean.

The area under a normal curve within one standard deviation of the mean is calculated below using the `pnorm()` function, which calculates the cumulative density function for a normal curve.

```{r}
stdDeviations <- 1

pnorm(stdDeviations) - pnorm(stdDeviations * -1)
```

The area under a normal curve within one standard deviation of the mean is depicted in @fig-zScoreDensity1SD.

```{r}
#| fig.cap: "Density of Standard Normal Distribution. The blue region represents the area within one standard deviation of the mean."
#| fig.scap: "Density of Standard Normal Distribution: One Standard Deviation of the Mean."
#| label: fig-zScoreDensity1SD
#| code-fold: true

x <- seq(-4, 4, length = 200)
y <- dnorm(x, mean = 0, sd = 1)
plot(x, y, type = "l",
     xlab = "z Score",
     ylab = "Normal Density")

x <- seq(stdDeviations * -1, stdDeviations, length = 100)
y <- dnorm(x, mean = 0, sd = 1)
polygon(c(stdDeviations * -1, x, stdDeviations),
        c(0, y, 0),
        col = "blue")
```

The area under a normal curve within two standard deviations of the mean is calculated below:

```{r}
stdDeviations <- 2

pnorm(stdDeviations) - pnorm(stdDeviations * -1)
```

The area under a normal curve within two standard deviations of the mean is depicted in @fig-zScoreDensity2SD.

```{r}
#| fig.cap: "Density of Standard Normal Distribution. The blue region represents the area within two standard deviations of the mean."
#| fig.scap: "Density of Standard Normal Distribution: Two Standard Deviations of the Mean."
#| label: fig-zScoreDensity2SD
#| code-fold: true

x <- seq(-4, 4, length = 200)
y <- dnorm(x, mean = 0, sd = 1)
plot(x, y, type = "l",
     xlab = "z Score",
     ylab = "Normal Density")

x <- seq(stdDeviations * -1, stdDeviations, length = 100)
y <- dnorm(x, mean = 0, sd = 1)
polygon(c(stdDeviations * -1, x, stdDeviations),
        c(0, y, 0),
        col = "blue")
```

The area under a normal curve within three standard deviations of the mean is calculated below:

```{r}
stdDeviations <- 3

pnorm(stdDeviations) - pnorm(stdDeviations * -1)
```

The area under a normal curve within three standard deviations of the mean is depicted in @fig-zScoreDensity3SD.

```{r}
#| fig.cap: "Density of Standard Normal Distribution. The blue region represents the area within three standard deviations of the mean."
#| fig.scap: "Density of Standard Normal Distribution: Three Standard Deviations of the Mean."
#| label: fig-zScoreDensity3SD
#| code-fold: true

x <- seq(-4, 4, length = 200)
y <- dnorm(x, mean = 0, sd = 1)
plot(x, y, type = "l",
     xlab = "z Score",
     ylab = "Normal Density")

x <- seq(stdDeviations * -1, stdDeviations, length = 100)
y <- dnorm(x, mean = 0, sd = 1)
polygon(c(stdDeviations * -1, x, stdDeviations),
        c(0, y, 0),
        col = "blue")
```

If you want to determine the *z* score associated with a particular percentile in a normal distribution, you can use the `qnorm()` function.
For instance, the *z* score associated with the 37th percentile is:

```{r}
qnorm(.37)
```

@@

Regression with a single predictor takes the form in @eq-regression.
A regression line is depicted in @fig-regression.
Multiple regression (i.e., regression with multiple predictors) takes the form in @eq-multipleRegression.

### 2024-06-26 multiple-regression.qmd
20240625 - multiple regression overview
20240626 - scores, scales, and regression

@@

- $B$ = unstandardized coefficient: direction and magnitude of the estimate (original scale)
- $\beta$ = standardized coefficient: direction and magnitude of the estimate (standard deviation scale)
- $SE$ = standard error: uncertainty of unstandardized estimate

>>>>

Regression with one predictor takes the form of @eq-regression:

$$
y = \beta_0 + \beta_1x_1 + \epsilon
$$ {#eq-regression}

where $y$ is the [outcome variable](#sec-correlationalStudy), $\beta_0$ is the intercept, $\beta_1$ is the slope, $x_1$ is the [predictor variable](#sec-correlationalStudy), and $\epsilon$ is the error term.

A regression line is depicted in @fig-regression.

```{r}
#| include: false
#| eval: false

set.seed(52242)
regression <- data.frame(outcome = rnorm(40, mean = 5, sd = 2))

regression$predictor <- complement(regression$outcome, .5)
regression$predictor <- regression$predictor + abs(min(regression$predictor))

lm(
  outcome ~ predictor,
  data = regression)

ggplot2::ggplot(
  data = regression,
  ggplot2::aes(
    x = predictor,
    y = outcome,
  )
) +
  ggplot2::geom_point() +
  ggplot2::geom_smooth(
    method = "lm",
    linewidth = 2,
    se = FALSE,
    fullrange = TRUE) +
  ggplot2::scale_x_continuous(
    lim = c(0,8),
    breaks = seq(from = 0, to = 8, by = 2),
    expand = c(0,0)
  ) +
  ggplot2::scale_y_continuous(
    lim = c(0,8),
    breaks = seq(from = 0, to = 8, by = 2),
    expand = c(0,0)
  ) +
  ggplot2::labs(
    x = "Predictor Variable",
    y = "Outcome Variable",
    title = "Regression Best-Fit Line"
  ) +
  ggplot2::theme_classic(
    base_size = 16) +
  ggplot2::theme(legend.title = element_blank())

ggsave("./images/regression.pdf", width = 6, height = 6)
```

::: {#fig-regression}
![](images/regression.png)

A Regression Best-Fit Line.
:::

Regression with multiple predictors—i.e., multiple regression—takes the form of @eq-multipleRegression:

$$
y = \beta_0 + \beta_1x_1 + \beta_2x_2 + ... + \beta_px_p + \epsilon
$$ {#eq-multipleRegression}

where $p$ is the number of [predictor variables](#sec-correlationalStudy).

@@

- $B$ = unstandardized coefficient: direction and magnitude of the estimate (original scale)
- $\beta$ (beta) = standardized coefficient: direction and magnitude of the estimate (standard deviation scale)
- $SE$ = standard error: uncertainty of unstandardized estimate

The unstandardized regression coefficient ($B$) is interpreted such that, for every unit change in the [predictor variable](#sec-correlationalStudy), there is a __ unit change in the outcome variable.
For instance, when examining the association between age and fantasy points, if the unstandardized regression coefficient is 2.3, players score on average 2.3 more points for each additional year of age.
(In reality, we might expect a nonlinear, inverted-U-shaped association between age and fantasy points such that players tend to reach their peak in the middle of their careers.)
Unstandardized regression coefficients are tied to the metric of the raw data.
Thus, a large unstandardized regression coefficient for two variables may mean completely different things.
Holding the strength of the association constant, you tend to see larger unstandardized regression coefficients for variables with smaller units and smaller unstandardized regression coefficients for variables with larger units.

Standardized regression coefficients can be obtained by standardizing the variables to [*z*-scores](#sec-zScores) so they all have a mean of zero and standard deviation of one.
The standardized regression coefficient ($\beta$) is interpreted such that, for every standard deviation change in the [predictor variable](#sec-correlationalStudy), there is a __ standard deviation change in the outcome variable.
For instance, when examining the association between age and fantasy points, if the standardized regression coefficient is 0.1, players score on average 0.1 standard deviation more points for each additional standard deviation of their year of age.
Standardized regression coefficients—though not the case in all instances—tend to fall between [−1, 1].
Thus, standardized regression coefficients tend to be more comparable across variables and models compared to unstandardized regression coefficients.
In this way, standardized regression coefficients provide a meaningful index of [effect size](#sec-practicalSignificance).
